---
title: "Make tidy data"
author: "Sten de Schrijver"
date: "2024-05-15"
output: html_document
---

The code in this file allows the creation of datasets in the tidy format for the optimal usage of the library Tidyverse in R. The two sections show the creation of such a dataset for the P1a and P1b datasets (see README for more information). These datasets have been received in a format comparable to the "dummy_data_P1a.xlsx" and the "dummy_data_P1b.xlsx". These Excel files have been filled with randomly generated data and do not reflect the actual data included in this study. Note that the comments for a line of code are given above the code. Further note that upon downloading this document, one could compile an HTML-file through the "Knit --> Knit to HTML" option.


## 1. Installing libraries and importing data
```{r setup}
# Specify vector with all required packages
list_of_packages <- c("tidyverse","readxl")
# Silence warnings and messages when importing
suppressWarnings(suppressMessages({
  # Make list of packages to install if not installed
  packages_to_install  <- list_of_packages[!(list_of_packages %in% installed.packages()[,"Package"])]
  for (lib in list_of_packages){
    # Install those packages that are not yet installed
    if (lib %in% packages_to_install){install.packages(lib)}
    # Load package into R session
    library(lib, character.only=TRUE)}
}))

# Load P1a data --> background data
dataset_P1a   <- read_excel("dummydata_background.xlsx",.name_repair = "unique_quiet") 
# Load P1b data --> epidemiological and clinical data
dataset_P1b_samples   <- read_excel("dummydata_P1b_samples.xlsx",.name_repair = "unique_quiet") 
# List of EV types and species
Enteroviruses         = read_excel("Enteroviruses.xlsx")%>% mutate(Species = as.factor(Species)) 
Viruses               = Enteroviruses$Virus                 
Species               = Enteroviruses$Species    
# The nomenclature on EVs has been updated since the data was requested
Nomenclature          = Enteroviruses$Newname     
# List of countries and regions
Regions               = read_excel("European_countries.xlsx") %>%
  mutate(across(c(Country,Region), as_factor))
```

## 2. Creating tidy data from P1a
```{r}
# The name of the P1a tidy dataset to be
P1a_data_name = "P1a_tidy.txt"
# Initiate a function to create the tidy data
Create_tidytable <- function(df){                                 
  # first row is empty, second row is the column-names
  N_files_P1A    = nrow(df)-2 
  # I initiate an empty dataset with 7 columns and 6 rows for every lab.  
  tidydf     = data.frame(matrix(ncol = 7, nrow = 6*N_files_P1A))   
  colnames(tidydf) = c("Country","Lab","Year","Tests","Positives","Typed_attempted","Typed_succesful") 
  # The data is from 2018-2023 (6 years)
  for (yy in 1:6){
    # Columns 3,4,5,6 for 2023, 7,8,9,10 for 2022, etc. With yy being 1:6 the right indexing is (4*yy)-1 : (4*yy)+2
    year_selection = df[,c((4*yy-1):(4*yy+2))]          
    # Within this selection, the year is specified in the first column
    year = year_selection[1,1]                        
    # In the former dataset the years have seperate columns. In the tidy dataset, the years will be in the same column, specified by a single variable, namely 'Year'. At every year, there will be a row for each institute with the proper data. Starting at row 0 until row 24 for 2023. This comes down to rows (yy-1)*N_files_P1A+1 (1 for 2023, 25 for 2022, etc.). 
    
    # First row per year
    starting_row = (yy-1)*N_files_P1A+1
    # Final row per year
    final_row    = yy*N_files_P1A
    # Filling in the empty columns with the right data
    tidydf[c(starting_row:final_row), c(4:7)] = year_selection[c(3:nrow(df)),]   
    # Filling empty column with the year
    tidydf[c(starting_row:final_row),3] = year                                   
    # Filling empty columns country and Lab
    tidydf[c(starting_row:final_row),c(1,2)]= df[c(3:nrow(df)),c(1,2)]         
    }
  tidydf = tidydf %>% arrange(Country,Lab,Year)
  return(tidydf)}
# Calling the function
Tidy_P1a = Create_tidytable(dataset_P1a) 
# displays the first 10 rows of the dataset
head(Tidy_P1a)
# Saving the data as comma delimited file with a .txt extension. Note that the delimiter can be changed to an character (i.e. semicolon, tab).
write.table(Tidy_P1a,file=P1a_data_name,sep=",",row.names = F)
```

## 3. Creating tidy data from P1b
```{r}
# Specify name of the output
P1b_data_name = "P1b_sample_tidy.txt"
# All different institutions 
Filecodes             = levels(as.factor(dataset_P1b_samples$File))
# Number of all different institutions
N_files = length(Filecodes)
# Virus nomenclature as named in the received dataset
Virs             = Enteroviruses$Virus
# New nomenclature
Newvirs          = Enteroviruses$Newname
# First column is nonsense
pb_sa                                           = dataset_P1b_samples[-1,]               
# Number of rows per lab (number of possible values for sample + a row of total numbers + row with virus names)
Length_sample_table                             = 9                             
# Change the possible values for sample, increasing readability
pb_sa[pb_sa=="BIOPSY/TISSUE"]                   = "Biopsy"                      
pb_sa[pb_sa=="BLOOD" ]                          = "Blood"                       
pb_sa[pb_sa=="FAECAL (stool, rectal swab)"    ] = "Faecal"                     
pb_sa[pb_sa=="RESPIRATORY (pharyngeal swab, nose swab etc.)" ] = "Respiratory" 
pb_sa[pb_sa=="UNKNOWN OR UNDEFINED" ]           = "Unknown"
pb_sa[pb_sa=="VESICLE/ SKIN SWAB"   ]           = "Vesicle"
pb_sa[pb_sa=="TOTAL NUMBER"]                    = "Total number"
# Currently there are many empty cells in the columns of country and lab (only filled for the first row per lab). This fills the complete columns
for (row in 1:N_files){
    # Select the row 'Total number' for every lab. this is the first row with an empty 'country' and 'file' columns.
    start_row = Length_sample_table*(row-1)+2
    # Select the final row that belongs to this lab
    final_row = row*Length_sample_table
    # For columns 1 (country) and 2 (file (=lab)), fill in the value in the first row for this lab
    pb_sa[start_row:final_row,c(1,2)] = pb_sa[(Length_sample_table*(row-1)+1),c(1,2)]
    } 
# The data for 2023 is in columns 4:13. For year 2022 this is 14:23, 2021 24:33, etc. Further, all columns 1:3 are added because these are the country, lab, and sample type
pb_sa_2023 = pb_sa[,1:13]                                                     
pb_sa_2022 = pb_sa[,c(1:3,14:23)]
pb_sa_2021 = pb_sa[,c(1:3,24:33)]
pb_sa_2020 = pb_sa[,c(1:3,34:43)]
pb_sa_2019 = pb_sa[,c(1:3,44:53)]
pb_sa_2018 = pb_sa[,c(1:3,54:63)]

# Now this is the datafiles we start with
head(pb_sa_2023)
```

### Freq-table function
```{r}
# Creates a frequency table, with each virus type in a different column. Input is: pb_dataset, the year, the new column (Month, Clinical signs, Age, and in this case, specimen type (sample))
# This function will be run for every year (2018-2023) separately
freq_table <- function(pb_data, yr, coln, L){
  # Number of columns of this table will be the number of viruses + year, country, lab and type of sample
  # Number of rows of this table will be equal to the number of labs times the number of rows per lab - 1 row (not interested in the row with EV types)
  table            = data.frame(matrix(ncol = length(Virs)+4,                   
                                       nrow = length(Filecodes)*(L-1)))         
  colnames(table)  = c("Country","Lab","Year",coln,Newvirs)
  # Copying the rows with the same order in sample type per lab per country, every 9th row starting from 1 (i.e. 1, 10, 19, 28, etc.) should be removed (this is the row with virus types)
  EVtype_rows = (L)*(1:N_files)-(L-1)
  table[,c(1,2,4)] = as.matrix(setNames(pb_data[-EVtype_rows,c(1,2,3)],NULL))
  # The year is inputted in the function
  table[,3]        = yr                 
  # Looping over all possible virus names
  for (virus_index in 1:length(Virs)){                                          
    EV = Virs[virus_index]               
    # Find all matches to this virus name
    Match_coordinates  = which(pb_data==EV, arr.ind=TRUE)
    # If there is a hit at all
    if (length(Match_coordinates)>0){         
      # Loop over the rows and columns listed in matching cells
      for (index in 1:nrow(Match_coordinates)){                                
        # Save the row number of the match in y
        y         = Match_coordinates[index,1]    
        # Save the column number of the match in x
        x         = Match_coordinates[index,2]       
        # Find the lab of of the match
        Labnumber = pb_data$File[y] 
        # For this specific match [y, x], the new table with designated virus (column) and lab, will be filled with the numbers that are listed underneath the virus type in the original dataset
        Case_numbers_to_copy = pb_data[y+(1:(L-1)),x]
        table[which(table$Lab==Labnumber),virus_index+4]=Case_numbers_to_copy
        }
      }     
    }
  # If the virus is not found, it will now have NAs for the labs. This turns those values to 0.
  table[is.na(table)]= 0  
  # Change all columns of the viruses to numeric rather than character
  table= table %>%                                                              
    mutate_at(c(5:ncol(table)), as.numeric)                                
  return (table)}
```

```{r}
# Running the freq_table function for all datasets per year
pb_sa_2023 = freq_table(pb_sa_2023,"2023","Sample",Length_sample_table) 
pb_sa_2022 = freq_table(pb_sa_2022,"2022","Sample",Length_sample_table)
pb_sa_2021 = freq_table(pb_sa_2021,"2021","Sample",Length_sample_table)
pb_sa_2020 = freq_table(pb_sa_2020,"2020","Sample",Length_sample_table)
pb_sa_2019 = freq_table(pb_sa_2019,"2019","Sample",Length_sample_table)
pb_sa_2018 = freq_table(pb_sa_2018,"2018","Sample",Length_sample_table)

# Display one of these tables
head(pb_sa_2023)
```

### Tidy-up function
```{r}
# This is a new function, that creates a tidy table from pb_sa_yyyy. An output is returned for each dataframe, all of which are merged in the end.
Tidy_up  <- function(df,newcol){  
  # Being ample to call a new column the string that the variable "newcol" encodes
  newcol = as.symbol(newcol)
  # Making a tidy table through pivoting longer. The viruses will be in a new column called "Virus" and the associated case numbers will be in a new column "Cases".
  long_df    = df %>%                                                 
     pivot_longer(!c(Country,Lab,Year,newcol),
                  values_to="Cases",                                            
                  names_to="Virus")%>%                                         
    # Initiate new and empty columns for the species (EV-A:D) and European region
    mutate(Species="") %>%                                                      
    mutate(Region="")                                                          

  # For sp = A:D and PEV
  for (sp in levels(Enteroviruses$Species)){                
    # Filter all possible EV types within this species
    Enteroviruses_sp       =   Enteroviruses%>% filter(Species==sp)       
    # At every position in the long dataframe where the virus type is in this "Enteroviruses_sp" list, the 7th column (Species) in this long_df will change to the correct species name
    long_df[(which(long_df$Virus %in% Enteroviruses_sp$Newname)),7]=sp}                   
    
  for (reg in levels(Regions$Region)){                                             
    # Filter all possible countries within this region
    countries_regs  =   Regions%>% filter(Region==reg)   
    # At every position in the long dataframe where the country is in this "countries_regs" list, the 8th column (Region) in this long_df will change to the correct Region name
    long_df[(which(long_df$Country %in% countries_regs$Country)),8]=reg}

  long_df$Cases   =  as.numeric(long_df$Cases)    
  long_df = long_df%>%
    mutate(Cases = as.numeric(Cases))%>%
    mutate(across(c(Country,Lab,Year,Virus,Species,Region,newcol), as_factor))

  long_df = long_df%>%arrange(Country,Lab,Year)
  return (long_df)}
```

```{r}
# Tidy sample data
pb_sa_2023_long = Tidy_up(pb_sa_2023,"Sample")
pb_sa_2022_long = Tidy_up(pb_sa_2022,"Sample")
pb_sa_2021_long = Tidy_up(pb_sa_2021,"Sample")
pb_sa_2020_long = Tidy_up(pb_sa_2020,"Sample")
pb_sa_2019_long = Tidy_up(pb_sa_2019,"Sample")
pb_sa_2018_long = Tidy_up(pb_sa_2018,"Sample")

# Display one of these tables
head(pb_sa_2023_long)
# These tables can be merged into one long format
Tidy_P1b_sample = rbind(pb_sa_2023_long,pb_sa_2022_long,pb_sa_2021_long,pb_sa_2020_long,pb_sa_2019_long,pb_sa_2018_long)

write.table(Tidy_P1b_sample,file=P1b_data_name,sep=",",row.names = F)
```
